---
title: "ESM 244 Lab 1"
author: "Joseph Kilgallen"
date: "1/6/2022"
output: html_document
---

```{r setup, include=TRUE, warning =FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(tidyverse)
library(here)
library(sf)
library(tmap)
```

### Read in the Data
```{r}
sf_trees <- read_csv(here('data', 'sf_trees', 'sf_trees.csv'),
                     show_col_types = FALSE)
```

### Part 1: Wrangling and ggplot review

**Example 1:** Find Counts of observation `legal_status` & wrangle a bit.
```{r}
### method 1: group_by() %>% summarize()
sf_trees %>% 
  group_by(legal_status) %>% 
  summarize(tree_count = n())

## method 2: different way plus a few new functions
top_5_status <- sf_trees %>% 
  count(legal_status) %>% # pretty much the same as using group_by and summarize
  drop_na(legal_status) %>% 
  rename(tree_count = n) %>% 
  relocate(tree_count) %>% 
  slice_max(tree_count, n= 5) %>%#slice_max() tells it to only take the top variables by within column specified and how many you want to keep i.e. n= X
  arrange(desc(tree_count)) #arrange reorganizes data, either high to low or opposite, desc() specifies high to low

```

#### Make graph of the top 5 from above:
```{r}
ggplot(data= top_5_status, aes(x= fct_reorder(legal_status, tree_count), y= tree_count)) +
  geom_col(fill= 'darkgreen') +
  labs(x= 'Legal Status', y= 'Tree Count', title = 'Plot of Tree Counts by Legal Status') +
  coord_flip() +
  theme_minimal()
```

**Example 2:** Only going to keep observations where legal status is "Permitted Site" and caretaker is "MTA", and store as `permitted_data_df` 

shift-cmd-c to comment/uncomment quickly
```{r}
# sf_trees$legal_status %>% unique()
permitted_data_df <- sf_trees %>% 
  filter(legal_status == 'Permitted Site', caretaker == 'MTA')
```

**Example 3** Only keep Blackwood Acacia Trees, and then only keep columns `legal_status`, `date`, `latitude`, `longitude`, and store as `blackwood_acacia_df`
```{r}
blackwood_acacia_df <- sf_trees %>% 
  filter(str_detect(species, 'Blackwood Acacia')) %>% ##str_detect tells it to look into the column specified i.e. "species" and then find the observations that match the description i.e. "Blackwood Acacia"
  select(legal_status, date, lat = latitude, long =longitude)

## Make a little graph of locations
ggplot(data= blackwood_acacia_df, aes(x= long, y= lat)) +
  geom_point(color= 'darkgreen') 

```

**Example 4** Use `tidyr::separate()` 
```{r}
## Separate function helps to take apart two parts of a column by a specifying where you want to separate from, in this instance we use ' :: ' to tell it to break apart the scientific and common names by finding the ::
sf_trees_sep <- sf_trees %>% 
  separate(species, into = c('spp_scientific', 'spp_common'), sep = ' :: ')
```

**Example 5** use `tidyr::unite() allows you to combine two columns
```{r}
ex_5 <- sf_trees %>% 
  unite('id_status', tree_id, legal_status, sep = '_cool_')
```

## Part 2: Make some maps
```{r}
###retrying the config with git
```


